Working title: Ecological network control

### 2018-04-28 paper/manuscript.Rmd
Ad syntheis heading to abstract
Add manuscript and supplementary information

---
title: "Quantifying the manageability of pollination networks in an invasion context"
classoption: a4paper
csl: journal-of-ecology.csl
output:
  pdf_document:
    fig_crop: no
  html_document: default
  word_document: default
header-includes:
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{caption}
- \usepackage{lineno}
documentclass: artikel1
bibliography: references.bib
---

@@

```{r libraries, echo = F, include = F}
library(igraph)
library(magrittr)
library(dplyr)
library(latex2exp)
library(ggplot2)
library(glmulti)
library(lme4)
library(grid)

"../code/functions" %>% 
	list.files(full.names = T) %>%
	plyr::l_ply(source)

"../code/plotting_functions" %>% 
	list.files(full.names = T) %>%
	plyr::l_ply(source)

```

```{r load data, include = F}
net <- "../data/processed/networks" %>%
	read_networks()
meta <- readr::read_csv("../data/ntw_info.csv") %>% dplyr::tbl_df() %>%
	dplyr::rename(count = method)
n_driver <- readRDS("../data/processed/n_driver.rds")
manag_models <- readRDS("../data/processed/models_manageability.rds")
src_bi_w <- readRDS("../data/processed/manageability_bi_vs_weight.rds")
r_w <- readRDS("../data/processed/random_weights.rds") %>% dplyr::tbl_df() %>%
  dplyr::rename_("net_name" = ".id")
r_d <- readRDS("../data/processed/random_directions.rds") %>% dplyr::tbl_df() %>%
  dplyr::rename_("net_name" = ".id")

prop <- expand.grid(c("asymmetry", "max_dep"), 
                    c("scaled_FALSE", "scaled_TRUE")) %>% plyr::alply(1)
driver <- c("../data/processed/matching_frequency") %>% 
  list.files(full.names = T) %>%
  lapply(list.files, full.names = T) %>% 
  unlist() %>% 
  lapply(read_allRDS)
driver <- mapply(function(x, y){
  attr(x, "type") <- as.character(y[,1])
  attr(x, "scaled") <- stringr::str_split(as.character(y[,2]), "_") %>% unlist() %>% extract(2) %>% as.logical()
  return(x)
}, driver, prop)
driver_bi <- "../data/processed/matching_frequency_bi" %>% 
  read_allRDS()
mod <- readRDS("../data/processed/detailed_species_models.rds")
mo <- readRDS("../data/processed/detailed_species_models_data.rds")
```

```{r plot theme, echo = FALSE, warning = F, message = F}
fer_theme <- theme_bw() +
	theme(text = element_text(family = "Helvetica"),
	      title = element_text(size = 7, hjust = 0),
	      legend.title = element_text(size = 8),
	      legend.text = element_text(size = 7),
				axis.text = element_text(size = 7),
				axis.title = element_text(size = 8, hjust = 0.5), 
				strip.text = element_text(size = 8, hjust = 0), 
				strip.background = element_blank(),
				plot.margin = grid::unit(c(5, 0.5, 2, 0), "mm"),
				panel.grid = element_blank())
```

```{r example networks, include = F, message = F, warning = F, }

sp_names_fancy <- c(TeX("$p_1$"), TeX("$p_2$"), TeX("$p_3$"), TeX("$a_1$"), TeX("$a_2$"))
sp_names_fancy_b <- c(TeX("$\\mathbf{p_1}$"), TeX("$\\mathbf{p_2}$"), TeX("$\\mathbf{p_3}$"), TeX("$\\mathbf{a_1}$"), TeX("$\\mathbf{a_2}$"))

sp_names <- c("p1", "p2", "p3", "a1", "a2")

adj_matrix <- c(0,0,0,1,0,
								0,0,0,2,5,
								0,0,0,0,0,
								0,0,1,0,0,
								0,0,1,0,0) %>%
	matrix(nrow = 5, byrow = T)

example_net <- adj_matrix %>% 
	graph_from_adjacency_matrix(mode = "undirected", weighted = TRUE) 
igraph::V(example_net)$type <- c(F, F, F, T, T)
igraph::V(example_net)$name <- sp_names

base_net <- example_net
bi_net <- example_net %>%
	keep_largest_component() %>%
	bipartite_digraph(type = "z-bi", weight.type = "asymetry") 
as_net <- example_net %>%
	keep_largest_component() %>%
	bipartite_digraph(type = "weight", weight.type = "asymmetry")

fam <- "Helvetica"

# layout is based on the bidirectional network
set.seed(1)
bi_net_uw <- bi_net
E(bi_net_uw)$weight <- 1
lay <- bi_net_uw %>%
	layout_nicely()
v_c <- "white"
l_c <- "black"
l_s <- 0.9
ma <- c(0,0,0,0)
label_pos_x <- -1
label_pos_y <- 1

pal <- RColorBrewer::brewer.pal(5, "PuOr")
```

@@

# Abstract

1. Despite important advances in network ecology, the link between the structure of an interaction network and management or conservation is still tenuous. Perhaps in part because of a lack of theoretical framework, efforts to strengthen this link have been mainly focused on whole-network properties instead of underlying mechanisms. 
2. Here, we leverage recent advances in control theory to develop an approach to assess the "manageability" of ecological networks. We then illustrate its application using ten pairs of uninvaded and invaded plant-pollinator communities. We also characterise species' suitability for inclusion in management interventions by exploring the entire space of alternative control strategies. Specifically, we measure the extent to which species (i) are necessary to steer the state of the community and (ii) are able to affect the abundance of other species. 
3. We found that the networks' manageability is most strongly determined by the ratio of plant to pollinator richness, which in turn constrains the networks' degree distribution. We also found that invasive plants have a dominant position in every invaded community and that this dominance is underpinned by high asymmetries in the dependences of their interaction partners. 
4. Our results provide novel insight into the design of ecologically-informed management interventions and lay the foundation for further work on the control of ecological networks.

**Keywords:** Network control theory, ecological networks, species importance, driver species, mutualism, management interventions

@@

In a complex system, the whole is often greater than the sum of its parts [@Montoya2006].
Within community ecology, a complex-systems approach has led to the development of analytical and simulation tools with which to understand, for example, the role of species embedded in a network of interactions [@Bascompte2009; @Stouffer2012]. 
The inherent complexity of nature, however, has regularly hindered—or at least complicated—our ability to find management solutions to many problems ecological communities face. 
To overcome this obstacle, we require a framework that allows us to explain, predict, and manage ecological communities, particularly when they are confronted with perturbations [@Green2005].
Ideally, such a framework is equipped to account for their complex structure and the dynamics that determine the species abundances and the state of the community.

Among the various possibilities, *control theory* appears to be a strong candidate [@Isbell2013]. Widely used in engineering to determine and supervise the behaviour of dynamical systems [@Motter2015], it is well equipped to deal with the many feedbacks present in ecological communities [@Liu2016]. 
Research in this area has established a strong link between the structure of complex networks and their *controllability*—the relative ability to manipulate network components to drive the system to a desired state [@Liu2011; @Cornelius2013; @Ruths2014]. 
These advances suggest that it is in principle possible to alter a whole ecological community's composition by modifying the abundances of only a few species.
Applications of control theory to ecological networks can also take into account the extent to which changes in the abundances of one species may ripple through the community [@Cornelius2013].
Therefore, control theory could also be harnessed to help identify which species are most relevant from a structural and dynamic perspective.

This information is valuable not only for basic ecology, but it might be also relevant to address more applied management and conservation challenges. 
This is particularly true in the context of biotic invasions, where identifying key players in the community is a prerequisite to informed attempts to alter the state of invaded ecosystems and maintain the state of uninvaded ones.
Despite recent advances in network theory, practical challenges to the conservation of interaction networks persist [@Tylianakis2010], and the link between the structure of complex networks and our ability to manage and conserve them is still ambiguous [@Bluthgen2010; @Kaiser-Bunbury2015].
To complicate things further, biotic invasions can induce dramatic changes in the patterns of interactions that determine the structure of ecological networks [@Tylianakis2008; @Ehrenfeld2010a], in particular pollination [@Olesen2002; @Aizen2008; @Bartomeus2008; @Vila2009; @Traveset2013]. 
Understanding how the differences in network structure before and after invasion impact our ability to manage the communities is thus a double challenge, but it is also the critical first step towards a fully informed recovery. 
Despite the apparent overlap, the control-theoretic perspective has not been adopted in an invasion context.

To bridge this gap, we outline an approach to apply control theory in an ecological context and implement it using empirical data.
Specifically, we use a set of ten pairs of uninvaded and invaded plant-pollinator communities to investigate the link between invasion, network structure and ecological management. 
While doing so, we focus on two particular questions.
First, grounded in the difficulties usually involved with invasive-species eradication and ecosystem restoration [@Woodford2016], we ask whether invaded networks have lower levels of "manageability" than their uninvaded counterparts; that is, whether they require a greater proportion of species to be managed to achieve the same level of control. 
Second, we ask whether some species are more important than others at driving the population dynamics of the community and which factors determine this importance.

# Theoretical framework

At the core of representing ecological communities as complex systems sits the idea that the state of any given node depends on its state, the state of the nodes it interacts with, the state of the nodes they interact with, and so on. 
These dependencies are given by the structure of the interactions, the dynamic relationships between nodes, and the mechanisms of self-regulation. 
This representation has been very useful for our understanding of ecological communities and might be advantageous for a more formal approach to ecological management based on control theory.

The overall objective of control theory is to be able to steer a system from one state to another in finite time. 
However, in complex systems like ecological communities, the large number of nodes and interactions as well as their non-linear dynamics render its control extremely challenging.
On one theoretical extreme, for instance, any ecological community could be fully controlled if we control the state of every species independently. 
On another, the mechanisms of self-regulation and the multiple feedback cycles found in ecological communities mean it is also theoretically possible to control the whole community by directly modifying the state of just a *single* species [@Cowan2012; @Markovsky2013]. 
Although mathematically correct, neither of these options are practical in real-world applications: the first because it is infeasible to design and implement interventions that modify the abundance of every single species in a community, and the second because the control signal might require unreasonably large amounts of time, energy, or unattainable rapid changes on the state [@Yan2012; @Motter2015].
Here we explore an intermediate point between these two trivial solutions.

In particular, we leverage the principle of species interdependence to find an intermediate set of *driver nodes* to control the network [@Liu2016]. 
When we focus on net interspecific effects, it is possible to identify a minimum number of nodes the control of which can theoretically drive the state of every other node in the network to a desired configuration. 
Conveniently the information necessary to determine this minimum number of driver nodes *D* is fully contained in the network structure [@Kalman1963; @Liu2011; @Motter2015]. 
Such a system can be described by $\frac{dx}{dt} = \mathbf{A}x + \mathbf{B}u(t)$, where the change of its state over time ($\frac{dx}{dt}$) depends on its current state *x* (for example, the species' abundances), an external time-varying input *u(t)* (the control signal), and two matrices *A* and *B*, which encode information about the network structure and how the species respond to the external input, respectively. 
If *S* is the number of species in the community, the matrix **A** has size $S \times S$ whereas the matrix **B** has size $S \times D$.

The goal of structural controllability, which we employ here, is to use the information contained in **A** to generate a supportable estimate of **B** (and by extension *D*).
This focus allows us to gain insight of the inherent controllability of a network, and the roles of the species that compose it, without being overly dependent on the particular choices of how the system dynamics are modelled or characterised. 
The trade-off of our approach is that, because of the assumption of linearity, structural controllability alone does not allow us to fully design the time-varying control signal *u(t)* that can drive the system from one particular equilibrium to another. 
Nevertheless, the lessons gained when assuming linearity---at both the network and the species level---are a prerequisite for eventually understanding nonlinear control [@Liu2011; @Liu2016].

## Manageability

The number of driver nodes *D* provides a structural indication of how difficult a network's control might be.
This is because systems that require a large number of external input signals are intuitively more difficult or costly to control.
In an ecological context, external inputs that modify the state of a node can be thought of as management interventions.
Therefore, the density of driver nodes $n_D = \frac{D}{S}$, where *S* is the total number of species in the community, is a measure of the extent to which network structure can be harnessed for network control.
For instance, a hypothetical "network" in which species do not interact would require direct interventions for every single species to achieve full control, whereas a linear food chain would require just one species to be directly controlled to harness cascading effects through its trophic levels. 
From this perspective, it is possible to use *n~D~* as an index of the manageability of an ecological community, understood in the context of how difficult is to modify the abundances of species in the community using external interventions—a common theme in ecosystem management, conservation, and restoration.

It has been recently shown that calculating *D* is equivalent to finding a maximum matching in the network [@Liu2011].
In a directed network, a *matching* consists of a subset of links in which no two of them share a common starting or ending node (Figure 1, Supporting Information S1).
A given matching has *maximal cardinality* if the number of matched links (also referred to as the *matching size*) is the largest possible.
A maximal cardinality matching is then called a *maximum matching* if the sum of the weights of the matched links (also referred to as *matching weight*) is again the largest possible [@West2001].

Once we have the subset of links that constitute a matching, we can also classify the nodes in the network based on that matching (Figure 1).
A node is called *matched* if it is at the end of a matched link and *unmatched* otherwise. 
A node is also called *superior* if it is at the start of a matched link.
Note that a node cannot be superior if it has no outgoing links.
Notably, these node categories are what helps us to link a maximum matching back to the concept of network controllability, as follows.
Unmatched nodes are the *driver nodes* *D* because they have no superior in the network and must be directly controlled by an external input [@Liu2011].
Each matched node, on the other hand, can be controlled by its superior.

```{r figure 1, echo = F, message = F, warning = F, fig.width = 6.2, fig.height = 1.9, results= "hide", fig.pos = "t", fig.cap = "**Matchings of a simple network**. (a) We start with a network in which the direction of the links indicates the potential direction of control; for example a link from *a~1~* to *p~1~* indicates that the state of *p~1~* is influenced by the state of *a~1~*. The numbers indicate the weight of each link. (b & c) This network has two maximal cardinality matchings; that is, two configurations in which it would be possible to exert full control of the network via external input signals to a minimal set of nodes. In both cases, the three matched links (purple arrows) represent the control paths through the network and provide an indication of the matched nodes (purple), which are controlled by superior nodes within the network (circular nodes). Unmatched nodes (orange) are called driver nodes because full network control requires external signals to be applied to them. Out of the two maximal cardinality matchings only one (c) has maximum weight and therefore is also a maximum matching. Further examples can be found in Supporting Information S1."}
layout(matrix(c(1, 1, 2, 4, 3, 4), 2, 3),
  	heights=c(8,2))
par( mar=c(0,0,0,0))

as_net %>% 
	plot(layout = lay, 
			 vertex.color = v_c, 
			 vertex.size = 35,
			 vertex.label = sp_names_fancy,
			 vertex.label.family = fam,
			 vertex.label.color = l_c,
			 edge.label = round(E(.)$weight,1), 
			 edge.label.family = fam,
			 edge.label.color = l_c,
			 edge.label.cex = l_s,
			 edge.curved = F,
			 edge.arrow.size = 0.5,
			 margin = ma)

text(label_pos_x, label_pos_y, "(a)", adj = c(0,0), font = 2)

as_mat <- as_net %>%
	digraph_bipartite() %>% 
	max_bipartite_match()

as_all_mat <- as_net %>%
	digraph_bipartite() %>%
	igraph::make_line_graph() %>%
	igraph::complementer() %>% 
	igraph::max_cliques(min = as_mat$matching_size) %>%
	rev()

text_pos_x <- 0.5
text_pos_y <- -0.7

base_as_net <- as_net
E(as_net)$weight <- round(E(as_net)$weight, 1)

1:length(as_all_mat) %>%
	plyr::l_ply(function(x){

		E(as_net)$matched <- 1
		E(as_net)$matched[as_all_mat[[x]]] <- 5

		E(as_net)$matched_width <- 1
		E(as_net)$matched_width[as_all_mat[[x]]] <- 1.5

		igraph::V(as_net)$matched <- 1
		igraph::V(as_net)$matched[igraph::V(as_net)$name %in%
																ends(as_net, as_all_mat[[x]])[, 2]] <- 5

		igraph::V(as_net)$matched_fill <- 2
		igraph::V(as_net)$matched_fill[igraph::V(as_net)$name %in%
																ends(as_net, as_all_mat[[x]])[, 2]] <- 4

		igraph::V(as_net)$superior <- 1
		igraph::V(as_net)$superior[igraph::V(as_net)$name %in%
																ends(as_net, as_all_mat[[x]])[, 1]] <- 2

		igraph::V(as_net)$sup_shape <- "square"
		igraph::V(as_net)$sup_shape[igraph::V(as_net)$name %in%
																ends(as_net, as_all_mat[[x]])[, 1]] <- "circle"

		names_fancy <- sp_names_fancy
		# names_fancy[igraph::V(as_net)$name %in%
		# 														ends(as_net, as_all_mat[[x]])[, 1]] <-
		# 	sp_names_fancy_b[igraph::V(as_net)$name %in%
		# 														ends(as_net, as_all_mat[[x]])[, 1]]

		matching_weight <- sum(E(as_net)$weight[as_all_mat[[x]]])

		as_net %>% 
			plot(layout = lay, 
					 # vertex.size = igraph::V(.)$superior * 10 + 20,
					 vertex.size = 30,
					 vertex.color = pal[igraph::V(.)$matched_fill],
					 vertex.shape = igraph::V(.)$sup_shape,
					 vertex.label = names_fancy,
					 vertex.label.family = fam,
					 vertex.label.color = l_c,
					 vertex.label.cex = 1,
					 # vertex.shape = igraph::V(.)$vertex.shape,
					 vertex.frame.color = pal[igraph::V(.)$matched],
					 edge.color = pal[E(.)$matched],
					 edge.label = NA, 
					 edge.width = E(.)$matched_width,
					 edge.label.family = fam,
					 edge.label.color = igraph::E(.)$matched,
					 edge.label.cex = l_s,
					 edge.curved = F,
					 edge.arrow.size = 0.5,
					 margin = ma) %>% 
			print()

		igraph::V(as_net)$name <- sp_names

		text(label_pos_x, label_pos_y, paste0("(",letters[1+x], ")"), adj = c(0,0), font = 2)
		text(text_pos_x, text_pos_y, cex = 0.9,
				 paste0("D = ", length(igraph::V(as_net)) - as_mat$matching_size, "\n",
				 			 "matching size: ", as_mat$matching_size, "\n",
				 			 "matching weight: ", round(matching_weight, 1)))
})

as_net <- base_as_net

plot.new()
legend(0.20, 0.8, legend = c("unmatched link", "matched link"), horiz = TRUE,
			 lty = 1, 
			 col = pal[c(1, 5)],
			 lwd = 2, cex = 0.9, xjust=0.5, yjust=0.8, bty = "n")

legend(0.835, 0.8, legend = c("unmatched node (driver)", "matched node"), horiz = TRUE,
			 pch = 21, 
			 pt.bg = pal[c(2, 4)],
			 col = pal[c(1,5)],
			 pt.cex = 2,
			 cex = 0.9, xjust=0.5, yjust=0.8, bty = "n")

legend(0.5, 0.35, legend = c("superior node", "non-superior node"), horiz = TRUE,
			 pch = c(21, 22), 
			 # pt.bg = pal[c(2, 4)],
			 pt.cex = 2,
			 cex = 0.9, xjust=0.5, yjust=0.8, bty = "n")

n
# 			 cex = 0.9, xjust=0.5, yjust=1, bty = "n")

```

Note that this framework requires a directed network in which the direction of the links corresponds to the direction of control. 
In the "Methods" section below, we explain our approach to determining the link direction in pollination networks.

## Relative importance

While calculating *n~D~* measures the manageability of an ecological community, it does not provide information about the identity of the species that compose the set of driver nodes. 
Ecologically, potential differences between species are relevant because management and conservation resources are limited, and therefore ecological interventions should be focused on the set of species that might provide the largest impact. 
Moreover, maximum matchings in a network are often not unique, and each maximum matching indicates unique paths that can potentially be used to control the network. 
We harness this property and use a network's complete set of maximum matchings to characterise each species' relative importance in driving the state of the community. 
One possibility is to characterise a species by the frequency *f~D~* with which it is classified as a driver node within this set of matchings.
However, the profile of our networks indicates that a large proportion of species were classified as driver nodes because of external interventions required to achieve **full** controllability and not because they influence the abundance of other species (Supporting Information S2).
Furthermore, the precise role of driver nodes is more ambiguous when full control is unfeasible or undesired—often the case in ecological settings.
We therefore also calculate the frequency *f~S~* with which a species is classified as a superior node since this is the frequency with which they form part of possible control paths.

Most commonly, structural controllability assumes unweighted networks—links exist or not, and hence *f~D~* and *f~S~* can be calculated by computing all possible maximum-cardinality matchings. 
However, we take the link weights into account when calculating the matchings here because the weights can reveal significant ecological patterns and processes that might be undetectable in unweighted networks [@Scotti2007; @Tylianakis2007; @Vazquez2007; @Kaiser-Bunbury2010a]. 
Additionally, species *A* may interact with both species *B* and *C* but depends strongly on *B* and only weakly on *C*. 
Intuitively, a management intervention designed to indirectly modify the abundance of species *A* is more likely to succeed if the abundance of *B*, rather than *C*, is directly controlled.
A complication of including the interaction weight when calculating the maximum matching, however, is that empirical interaction strengths are to some extent stochastic and depend on proximate factors such as sampling method and intensity [@Gibson2011].
We overcome this issue by calculating all maximal cardinality matchings and then ranking them by their matching weight. 
By following this approach, we effectively give priority to the species that participate in the pathways that potentially have the largest impact on the community while acknowledging the limitations associated with sampling and its potential restrictions [@Jordano2016].

@@

```{r network summaries, include = F}
lop <- filter(meta, study == "lopezaraiza") %>% mutate(size = n_pla + n_pol) %$% size %>% summary() 
bar <- filter(meta, study == "bartomeus") %>% mutate(size = n_pla + n_pol) %$% size %>% summary() 
```

We next applied the previously defined framework to ten paired pollination networks.
Each network pair was composed of a community invaded by a plant and a community "free" of the invasive species. 
Four pairs were obtained from natural or semi-natural vegetation communities in the city of Bristol, UK [@Lopezaraiza-Mikel2007]. 
These networks are comprised of `r lop[1]`--`r lop[6]` species (mean `r round(lop[4])`), and non-invaded plots were obtained by experimentally removing all the flowers of the invasive species *Impatients grandulifera*.
The other six pairs were obtained from lower diversity Mediterranean shrublands in Cap de Creus National Park, Spain [@Bartomeus2008].
These networks are comprised of `r bar[1]`--`r bar[6]` species (mean `r round(bar[4])`); in contrast to the above, uninvaded communities were obtained from plots that had not yet been colonised by either of the invasive species *Carpobrotus affine acinaciformis* or *Opuntia stricta*. 
Further details about the empirical networks can be found in Supporting Information S3.

We then specified the structure of all networks using pollinator visitation frequency, which has been shown to be an appropriate surrogate for interspecific effects in pollination networks [@Vazquez2005; @Bascompte2006]. 
To further examine whether this decision would influence our results, we also evaluated the effect of using pollinator efficiency or pollinator importance as alternative measures in a different data set that lacked invasive species [@NeEman2010; @Ballantyne2015], and we found quantitatively similar results for all three options (Supporting Information S4).
Because our approach depends on the network topology, we also evaluated the robustness of our results to the undersampling of interactions.
Specifically, we calculated *n~D~* and species relative importance for 500 random subsamples of each empirical network in which the weakest links were more likely to be removed. 
Our sensitivity analysis indicated that, even in the absence of complete sampling, a control-theoretic approach can still be applied (Supporting Information S5).

## Manageability

```{r component_lost, echo = F, eval = T}
a <- net %>%
	lapply(igraph::E) %>%
	lapply(length)

b <- net %>%
	lapply(keep_largest_component) %>%
	lapply(igraph::E) %>%
	lapply(length)

int_lost <- ((mapply(magrittr::subtract, a, b) / unlist(a)) * 100) %>% mean() %>% round(1)

a <- net %>%
	lapply(igraph::V) %>%
	lapply(length)

b <- net %>%
	lapply(keep_largest_component) %>%
	lapply(igraph::V) %>%
	lapply(length)

sp_lost <- ((mapply(magrittr::subtract, a, b) / unlist(a)) * 100) %>% mean() %>% round(1)
```

We began by quantifying the manageability of each network.
To do so, we calculated the networks' maximum matching and determined the minimum proportion of species *n~D~* that need external input signals to fully control the species abundances in the community. 
Note that because all maximum matchings have the same matching size, it is only necessary to calculate one of them. 
To simplify the analysis, if a network had more than one component (two species are in different components if there exists no path between them and are hence independent of each other in terms of network control) we only considered the largest. 
Smaller components were present in eleven out of the twenty networks and were typically composed of just one plant and one pollinator. 
Their removal represented an average loss of `r sp_lost`% of the species and `r int_lost`% of the interactions.

### Weighting & directing links

As we noted earlier, the maximum-matching approach requires a directed network in which a link from species *i* to species *j* indicates that the abundance of *j* can be affected by the abundance of *i*.
This implies that we need first to identify a directionality for the links between species that is consistent with the dynamics of the community (Figure 2). 
In some ecological networks, establishing the directionality can appear relatively straightforward, for example when links represent biomass transfer or energy flow [@Isbell2013]. 
Interspecific effects in pollination networks, however, are not strictly directed since the benefit is mutual between interacting species. 
Nevertheless, the relative extent to which a given pair of interacting species affect each other can be quantified by the magnitude of the mutual dependence between them [@Bascompte2006]. 
The dependence of plant *i* on pollinator *j*, *d~ij~*, is the proportion of the visits from pollinator *j* compared to all pollinator visits to plant *i*. 
Likewise, the dependence of pollinator *j* on plant *i*, *d~ji~*, is the proportion of the visits by pollinator *j* to plant *i* compared to all visits of pollinator *j*.
Using dependences generates a weighted bipartite network in which all interacting pairs are connected by two directed links (Figure 2b).

>>>>

---
title: "Working title: Ecological network control"
classoption: a4paper
csl: journal-of-ecology.csl
output: 
  bookdown::pdf_document2: 
    keep_tex: yes
    number_sections: false
toc: false
header-includes:
- \usepackage{geometry}
- \geometry{right=3in,marginparwidth=2.5in}
- \usepackage{setspace}
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{caption}
- \usepackage{lineno}
- \usepackage{xr}
- \externaldocument[S-]{supp-info}
documentclass: artikel1
bibliography: references.bib
---

@@

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, 
                      warning = FALSE)
options(xtable.comment = FALSE, 
        xtable.caption.placement = 'bottom',
        xtable.size = 'small',
        booktabs = TRUE)
```

@@

A major goal in ecological research is to understand the role that species play in the biotic environment.
Within community ecology, a complex-systems has led to the development of a variety analytical and simulation tools with which to understand the role of the role of species embedded in a network of interactions [@bascompte_assembly_2009; @stouffer_evolutionary_2012]. 
A particularly relevant dimension of any species' role is its ability to alter the abundance of other species in the community, since this impacts ecosystem function and biodiversity and might have knock down effects on other processes and services.
This ability is often referred as "keystoneness". 
A significant proportion of the network tools used to estimate species roles in this context relies on the assumption that there is a link between a species centrality—a relative ranking of its importance that stems from social network research—and its ability to alter other species' abundance. 
At first glance, this assumption makes sense. 
At the core of representing ecological communities as networks sits the idea that the abundance of a species also depends on the abundance of its neighbours (the species it interacts with) and, by extension, the abundance of its neighbours neighbours', and so on. 
More central species, by definition, are better connected and hence should be more likely to affect the abundance of other species in the network.

Though conceptually intuitive, the relationship between centrality and keystoness is largely phenomenological. 
On one hand, there is more than a dozen indices of centrality used in ecology. 
Chosing a particular metric can be a difficult task, especially because to date there are not clear methods to determine the suitability of an index to a particular application. 
Furthermore, recent studies have shown that definitions of keystoness based on centrality might not adequately predict the cascading extintions keystone species are ususally associated to. 
As a result, community ecology could benefit from an alternative, more mechanistically-grounded, approach to understand how species affect each other abbundance. 
Species abundances are also influenced by the dynamic relationships between nodes, and the mechanisms of self-regulation, however, community and population dynamics can be modeled in innumerable ways, and empirical support for one versus another is often still ambiguous. 
The alternative approach should acknowledge ecosystem dynamics, but without being overly dependent on the particular choices of how they are characterised.

Among the various possibilities, *control theory* appears to be a strong candidate [@isbell_human_2013]. 
Control theory is a widely studied branch of engineering used to determine and supervise the behaviour of dynamical systems [@motter_networkcontrology_2015]. 
It is inherently designed to deal with system feedbacks and its application has recently been expanded to complex networks [@liu_control_2016]. 
In correspondence with ecological questions, advances in this field have established a clear link between the structure of the network and the way nodes affect each other. 
Unlike centrality indices, however, this link is not based on a priori assumptions between network metrics and keystoness, instead on well-established advances in both dynamical and complex-systems theory <!--STOUFFER: Are we shooting ourselves in the foot since this isn't entirely true? That is, see comments from ELE about linearity and otherwise. CAGUA: How about this modified phrasing?-->.

Despite the apparent overlap, the concepts and approaches of control theory are not yet commonly applied in an ecological context. 
Possibly, this is in part because research on the control of complex networks is generally more concerned with the state of the system as a whole. 
This is in contrast with the ecological role concept, where the focus is on the nodes that compose the network.
At its fundamental level, control theory first determines whether a system is controllable or not; that is, it asks if a system can be driven to a desired state within a finite amount of time. 
Although the controllability of a network is a whole-system property, it has been recently shown that asking for the controllability of a complex-system is equivalent to finding a particular set of nodes. 
A set with which is possible to control the state of the whole network. 
Notably, this set of nodes is not always unique for a given network.
This implies that an examination of the distinct sets provides a means to connect nodes with its *general* ability to modify the system to which they belong.

Here, we apply methods from control theory to an ecological problem and show how it can be used to gain insight into the role of species in an ecological network. 
Specifically, we outline the approach using a set of ten pairs of uninvaded and invaded plant-pollinator communities. 
We use invaded communities because there is strong empirical evidence showing that invasive species play an important role shaping the abundances of other species, and particularly in these ten networks [@lopezaraizamikel_impact_2007; @bartomeus_contrasting_2008-1]. 
This choice thus offers us an opportunity to explicitly contrast theoretical observations with empirical evidence.
First, grounded in the difficulties usually involved with invasive-species eradication and ecosystem restoration [@woodford_confronting_2016], we ask whether there are differences between the controllability of invaded and uninvaded networks. <!--STOUFFER: I'd remove the grounded in.... stuff, but that's just me.  CAGUA: I don't like it either. I tried removing but then it feels like this question is coming out of the blue.-->
Then we expand existing methods control theory to effectively link the *controllability* of a network with the role of particular nodes. 
We ask—from a control-theoretic perspective—whether there are key differences between species in the role they play at driving the population of other members of the community and identify the ecological factors related to these differences. 
Finally, we compare the proposed approach to current methods based on species centrality and show how these methods are indeed valuable but only paint a limited picture in regards to the "keystoneness" of a species.

@@

```{r load-metadata}
drake::loadd(metadata)
lop <- dplyr::filter(metadata, study == "lopezaraiza") %>% 
  dplyr::mutate(size = n_pla + n_pol) %$% size %>% summary() 
bar <- dplyr::filter(metadata, study == "bartomeus") %>% 
  dplyr::mutate(size = n_pla + n_pol) %$% size %>% summary() 
```

We used ten paired pollination communities to test the control theoretic approach.
Each community pair was composed of a community invaded by a plant and a community free of the invasive species. 
Four pairs correspond to natural or semi-natural vegetation communities in the city of Bristol, UK [@lopezaraizamikel_impact_2007]. 
These communities are comprised of `r lop[1]`--`r lop[6]` species (mean `r round(lop[4])`), and non-invaded plots were obtained by experimentally removing all the flowers of the invasive species *Impatients grandulifera*.
The other six pairs were obtained from lower diversity Mediterranean shrublands in Cap de Creus National Park, Spain [@bartomeus_contrasting_2008-1].
These communities are comprised of `r bar[1]`--`r bar[6]` species (mean `r round(bar[4])`); in contrast to the above, uninvaded communities were obtained from plots that had not yet been colonised by either of the invasive species *Carpobrotus affine acinaciformis* or *Opuntia stricta*. The structure of all these communities was defined by the pollinator visitation frequency, which has been shown to be an appropriate surrogate for interspecific effects in pollination networks [@vazquez_interaction_2005; @bascompte_asymmetric_2006]. 
Full details about the empirical networks, the implications of using visitation frequency to quantify the strength of the associations, and the sensitivity of the control theoretic approach to undersampling of the interactions can be found in the Supplementary materials SX.

The first step before applying methods of control theory is to construct a directed network that can provide an indication of the extent to which species affect each others abundance. 
In some ecological networks, establishing the directionality can appear relatively straightforward, for example when links represent biomass transfer or energy flow [@isbell_human_2013]. 
In pollination networks, however, this directionallity is less obvious as both species can in theory benefit from the interaction. 
We overcome that obstacle by noting that the extent to which species *i* affects species *j* relative to the extent to which *j* affects *i* can be summarised by the interaction asymmetry [@bascompte_asymmetric_2006]. 
This asymmetry is given by $$a(i,j) = a(j,i) = \frac{\left | d_{ij}-d_{ji} \right |}{max\left ( d_{ij}, d_{ji} \right )},$$ where the dependence of plant *i* on pollinator *j*, *d~ij~*, is the proportion of the visits from pollinator *j* compared to all pollinator visits to plant *i*. 
Previous research has shown that mutualistic interactions are often highly asymmetric in natural communities; in other words, if a plant species is largely dependent on a pollinator species, then that pollinator tends to depend rather weakly on the plant (or vice versa). 
By creating a directed link based on the asymmetry of their dependence we are able to generate a network that is consistent with the dynamics of the community while satisfyingh the requirements of structural controllability.

After constructing the directed network, we can now calculate the controllability of the different networks and investigate wether there are differences between invaded and uninvaded communities.

## Controllability

A system is said to be controllable if it is possible to steer it from an initial to an arbitrary final state within finite time [@kalman_mathematical_1963]. 
A simple dynamic system can be described by $\frac{dx}{dt} = Ax + Bu(t)$, where the change of its state over time ($\frac{dx}{dt}$) depends on its current state *x* (for example, the species' abundances), an external time-varying input *u(t)* (the control signal), and two matrices *A* and *B*, which encode information about the network structure and how species respond to external inputs, respectively. 
In classic control theory, determining wether this system is controllable or not can easily be achieved by checking that its controllability matrix $R= [ \begin{matrix}B & AB & A^{2}B & ...& A^{n-1}B\end{matrix} ]$ has full rank. 
In complex systems, however, employing this rank condition, or numerical approximations of it, is unfeasible because it is hard to fully parameterize *A* and *B*. 
Here we use an approach based on structural controllability, which assumes that we are confident that the elements of *A* and *B* have either non-zero or zero values (there is an interaction or not), but that we less sure about the precise magitude of the non-zero values [@lin_structural_1974]. 
Essentially controllability is translated into a topological problem: a network is controllable if there are no inaccessible nodes—nodes without incoming links—or dilations—*expansions* of the network (Fig X).

We are often able to estimate *A* in ecological networks, as it represents the interactions between species. 
Part of the control problem resides on estimating a supportable estimation of *B*, which represents the links between external inputs and species. 
Naively, any ecological community (and any system for that matter) could be controlled if we control the state of every species independently. 
Such approach is typically unpractical so we are interested in finding a minimum driver node set (effectively finding *B*) with which to make the system contr

Controllability is this
But is hard to calculate.
Therefore we use recent advances that show that it can be calculated using maximum matching algorithms.

A maximum matching is this. 
And can be calculatid this way. 
And gives this information. 
But its not unique.
This can provide extra information. 
Therefore we calculate them all.
